An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-04 EconGrowthSeminar.Rnw
Now including a figure for the seminar. Lots more of the seminar outline in beamer has been developed.

  \setbeamercolor{background canvas}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block title}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{block body}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block title alerted}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block body alerted}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block title example}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block body example}{bg=CalvinLightTan!60!white}

@@

  \setbeamercolor{palette sidebar primary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{palette sidebar secondary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{palette sidebar tertiary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{palette sidebar quaternary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}

@@

\begin{frame}{Example Frame from the Template}{Shows how to overlay stuff} %----------------------------

  You can create overlays\dots
  \begin{itemize}
  \item using the \texttt{pause} command:
    \begin{itemize}
    \item
      First item.
      \pause
    \item    
      Second item.
    \end{itemize}
  \item
    using overlay specifications:
    \begin{itemize}
    \item<3->
      First item.
    \item<4->
      Second item.
    \end{itemize}
  \item
    using the general \texttt{uncover} command:
    \begin{itemize}
      \uncover<5->{\item
        First item.}
      \uncover<6->{\item
        Second item.}
    \end{itemize}
  \end{itemize}
\end{frame}

\subsection[Growth Modeling]{Economic Growth Modeling} %+++++++++++++++++++++++++++++++++++++++

\begin{frame}{Factors of Production: $K$ and $L$} %----------------------------
  \begin{itemize}
  \item
    Capital Stock ($K$) <Definition here>
  \item
    Labor ($L$) <Definition here>
  \item
    Energy ($E$)
  \end{itemize}
\end{frame}

\subsection{Types of Energy} %+++++++++++++++++++++++++++++++++++++++

\begin{frame}{Thermal Energy ($Q$)} %----------------------------

@@

\begin{frame}{Exergy ($X$)} %----------------------------

@@

\begin{frame}{Useful Work ($W$)} %----------------------------

@@

\subsection{Economic Growth Models} %+++++++++++++++++++++++++++++++++++++++

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sources of Data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection[Econ Data]{Sources of Economic Data} %+++++++++++++++++++++++++++++++++++++++

@@

@@

@@

@@

@@

\begin{frame}{Summary} %----------------------------

  % Keep the summary *very short*.
  \begin{itemize}
  \item
    The \alert{first main message} of your talk in one or two lines.
  \item
    The \alert{second main message} of your talk in one or two lines.
  \item
    Perhaps a \alert{third message}, but not more than that.
  \end{itemize}

  % The following outlook is optional.
  \vskip0pt plus.5fill
  \begin{itemize}
  \item
    Outlook
    \begin{itemize}
    \item
      Something you haven't solved.
    \item
      Something else you haven't solved.
    \end{itemize}
  \end{itemize}
\end{frame}

>>>>

  \setbeamertemplate{background canvas}[vertical shading][top=CalvinLightTan, bottom=white]
  \setbeamercolor{block title}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{block body}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block title alerted}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block body alerted}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block title example}{bg=CalvinLightTan!60!white}
  \setbeamercolor{block body example}{bg=CalvinLightTan!60!white}

@@

  \setbeamercolor{palette sidebar primary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{palette sidebar secondary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{palette sidebar tertiary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{palette sidebar quaternary}{fg=CalvinMaroon, bg=CalvinLightTan!60!white}
  \setbeamercolor{sidebar}{bg=CalvinLightTan}

@@

\begin{frame}{Energy is Essential for Economic Activity} %----------------------------
  What do you do that doesn't require energy?
\end{frame}

\begin{frame}{Cleveland, 1984} %----------------------------
  Show Cleveland's graph and cover of \emph{Science} here.
\end{frame}

\begin{frame}{Problem: Energy Cost Share is Low} %----------------------------
  Therefore, energy is considered to be economically unimportant.

  Compare cost shares of capital stock, labor, and energy.
\end{frame}

\begin{frame}{Energy Cost Share and Economic Growth} %----------------------------
  Show graphs of energy cost share and recessions
\end{frame}

\begin{frame}{The Big Questions} %----------------------------
  How account for energy in economic models?

  What happens when energy supplies become constrained?

  Historical economic development paths are energy intensive. What happens when energy is constrained and prices rise? Will developing countries be able to expand their economies?
\end{frame}

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure/Figure_Growth_Models.pdf}
  \end{center}
  \begin{itemize}
  \item
    Goal: predict economic output over time
  \item
    Economic output is a function of ``factors of production''
  \item
    Output defined as GDP ($Y$)
  \item
    All inputs and outputs normalized to initial year (0)
  \end{itemize}
\end{frame}

\subsection[Factors of Production]{Factors of Production} %+++++++++++++++++++++++++++++++++++++++

\begin{frame}{Capital Stock ($K$) and Labor ($L$)} %----------------------------
  \begin{itemize}
  \item
    Capital Stock ($K$) <Definition here>
  \item
    Labor ($L$) <Definition here>
  \end{itemize}
\end{frame}

\begin{frame}{Energy ($E$)} %----------------------------
  Caleb complete
\end{frame}

\begin{frame}{Thermal Energy ($Q$)} %----------------------------
  Caleb complete
\end{frame}

\begin{frame}{Exergy ($X$)} %----------------------------
  Caleb complete
\end{frame}

\begin{frame}{Useful Work ($W$)} %----------------------------
  Caleb complete
\end{frame}

\subsection[Growth Models]{Economic Growth Models} %+++++++++++++++++++++++++++++++++++++++

  \begin{equation*} \label{eq:Single_Factor_Generic}
    y = \mathrm{e}^{\lambda (t-t_0)}f^{m}
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $y$   & $Y$/$Y_0$ \\
      $t$   & year\\
      $t_0$ & initial year\\
      $f$   & $F$/$F_0$, a factor of production\\
      \hline
      $\lambda$   &   technological progress\\
      $m$   &         ``strength'' of factor $f$\\
      \hline
    \end{tabular}
  \end{center}

@@

\begin{frame}{Example Single-Factor Fit} %----------------------------
  Heun complete
\end{frame}

  \begin{equation*} \label{eq:CD_With_Energy}
    y = \mathrm{e}^{\lambda (t-t_0)}k^{\alpha}l^{\beta}e^{\gamma}
  \end{equation*}

  \begin{equation*} \label{eq:CD_Const_Returns_To_Scale}
    \alpha + \beta + \gamma = 1
  \end{equation*}

  \begin{equation*} \label{eq:CD_Constraints}
    0 \leq \alpha, \beta, \gamma \leq 1
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\lambda$   &   technological progress\\
      $\alpha$   &    ``strength'' of capital\\
      $\beta$   &    ``strength'' of labor\\
      $\gamma$   &    ``strength'' of energy\\
      \hline
    \end{tabular}
  \end{center}

@@

\begin{frame}{Example Cobb-Douglas Fit} %----------------------------
  Heun complete
\end{frame}

  \begin{equation*} \label{eq:micEconCES_model}
    \begin{split}
    y = & \gamma \mathrm{e}^{\lambda(t-t_0)} \\
        & \cdot \left\{\delta \left[\delta_1 k^{-\rho_1} + (1-\delta_1)l^{-\rho_1} \right]^{\rho/\rho_1} 
                        + (1-\delta) e^{-\rho} \right\}^{-1/\rho}
    \end{split}
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $\gamma$   &   fitting coefficient\\
      $\lambda$  &   technological progress\\
      $\delta$   &   split between ($kl$) and ($e$)\\
      $\rho$     &   ``strength'' of ($kl$) vs. ($e$)\\
      $\delta_1$   &   split between ($k$) and ($l$)\\
      $\rho_1$     &   ``strength'' of ($k$) vs. ($l$)\\
      \hline
    \end{tabular}
  \end{center}

@@

\begin{frame}{Example CES Fit} %----------------------------
  Heun complete
\end{frame}

  \begin{equation*} \label{eq:LINEX_model}
    y = e \: \mathrm{e}^{\left[a_0 \left(2 - \frac{l + e}{k} \right) + a_0 c_t \left(\frac{l}{e} - 1 \right)\right]}
  \end{equation*}

  \begin{center}
    \begin{tabular}{|c|l|}\hline
      $k$   & $K$/$K_0$, capital stock\\
      $l$   & $L$/$L_0$, labor\\
      $e$   & $E$/$E_0$, any of $q$, $x$, or $u$\\
      \hline
      $a_0$   &   fitting parameter\\
      $c_t$   &   fitting parameter\\
      \hline
    \end{tabular}
  \end{center}

@@

\begin{frame}{Example LINEX Fit} %----------------------------
  Heun complete
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Data Sources]{Sources of Data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection[Economic Data]{Sources of Economic Data} %+++++++++++++++++++++++++++++++++++++++

\begin{frame}{Economic Data}
  Lucas complete
\end{frame}

@@

\begin{frame}{Energy Data}
  Caleb complete
\end{frame}

@@

\begin{frame}{US Single Factor}
  Heun complete
\end{frame}

\begin{frame}{US Cobb-Douglas}
  Heun complete
\end{frame}

\begin{frame}{US CES}
  Heun complete
\end{frame}

\begin{frame}{US LINEX}
  Heun complete
\end{frame}

@@

\begin{frame}{ZA Single Factor}
  Heun complete
\end{frame}

\begin{frame}{ZA Cobb-Douglas}
  Heun complete
\end{frame}

\begin{frame}{ZA CES}
  Heun complete
\end{frame}

\begin{frame}{ZA LINEX}
  Heun complete
\end{frame}

@@

\begin{frame}{Results}
  Heun complete
\end{frame}

@@

\begin{frame}{Conclusions} %----------------------------

  \begin{itemize}
  \item
    Including energy is unwarranted \\
    \scriptsize
    Little statistical evidence that including energy in production functions improves fit.
    \normalsize
  \item
    Including energy is unnecessary \\
    \scriptsize
    Energy is always \emph{already} embedded in capital stock ($K$) and labor ($L$).
    \normalsize
  \vskip0pt plus.5fill
  \item
    Need a modeling method that captures energy's embeddedness \\
    \scriptsize
    System dynamics modeling
    \normalsize
  \end{itemize}
